Safe Memory Regions for Big Data Processing

### 2016-10-10 paper/type-inference.tex
type inference proofs

\name's region type system imposes a heavy annotation burden, and
manually annotating C\# standard libraries with region types
can be tedious. We now present our region type inference algorithm
that eliminates the need to write region type annotations.
% except on some higher-order functions.
Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.

@@

%\item
%\label{item:context}
%Consider any constraint of the form $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$,
%where $\pi_i$ and $\pi_j$ are both region constants.
%If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
%$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
%Otherwise, we check if $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.
%If this entailment does not hold, the constraint solver fails, since no valid solution is possible.
%
%\item
%    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
%$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
%If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
%solver fails.
%
%\item
%When a region variable $\rho$ is unified with a region constant $\pi$,
%if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
%and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
%If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.
%\item
%  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
%  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
%  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$. 
\end{enumerate}

Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.
% the special structure of the generated
% constraints: namely that $\phictxt$  is a conjunction of outlives-constraints of the
% form $\pi \outlives \pi_s$ where $\pi_s$ is a static region identifier and $\pi$ is a
% region constant. As a result, we can show that $\isvalid{\varphi \conj \phictxt}{\pi_i \outlives \pi_j}$
% iff $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ or $\isvalid{\phictxt}{\pi_i \outlives \pi_j}$.

@@

@@

Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.

\TODO{The following is an informal sketch, requiring some formalization.}

The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
It is easy to show that we take care of these.

The more difficult case are validity constraints, involving either region variables
($\rho$) or predicate variables ($\varphi$) or both.

The constraint-generation rules guarantee that every region variable will
be unified with some region constant\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value

We use this property to show that it is okay to ignore validity constraints
in $\satC$ containing region variables. For every
region variable $\rho$ unified with a constant $\pi$, and for every validity
constraint containing $\rho$, we will have a corresponding validity constraint
containing $\pi$ instead.
Note: We likely need to add another rule to the saturation rules for computing
$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
We need this to propagate constraints from an outer scope into an inner scope.

As for validity constraints containing a predicate variable $\varphi$ on the consequent,
we argue that the use of the Substitution rule in computing $\satC$ makes it okay
to ignore these constraints.

As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
us convert these into simpler constraints that do not have $\varphi$ on the
antecedent or into a constraint that is directly incorporated into definition
of $\predSubstFn$.
\end{proof}

@@

\begin{itemize}
  \item explain: non-determinism in defining region-variable binding is immaterial.
  \item explain lemma 2's use case (a) and (b)
  \item notation: decomposable is dependent on C
  \item cleanup: syntax of constraints (region-constraint, validity constraint, etc.)
    and later usage (abduction constraint)
  \item adjust spacing in RULEs
  \item for function-calls: need basic typing from core language to introduce region parameters
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item update abstract/intro to stress aspects of constraint solver
  \item Change static semantics to use method precondition, without class precondition
\end{itemize}

>>>>

\name's region type system imposes a heavy annotation burden, and
manually annotating C\# standard libraries with region types
can be tedious. We now present our region type inference algorithm
that eliminates the need to write region type annotations.
% except on some higher-order functions.
Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.
For simplicity, we assume that each \C{letregion} and \C{open} construct
in the input program introduces a distinct region identifier $\pi$.

@@

Point~\ref{item:context} makes use of Lemmas~\ref{lemma:gc-is-decomposable} and~\ref{lemma:decomposition}.

@@

Consider a set of constraints $D$. We say that a region variable $\rho$
\emph{occurs} in a \emph{context} $\ell$ (in $D$) if $D$ contains some
constraint $\isvalid{\ell}{r}$ where $\rho$ occurs in $r$.
We say that a region variable $\rho$ is \emph{bound} to a region constant
$\pi$ in a context $\ell$ if
$\{ \isvalid{\ell}{\rho \outlives \pi}, \isvalid{\ell}{\pi \outlives \rho} \} \subseteq D$.
We say that a region variable $\rho$ is \emph{bound} to a region constant $\pi$ (in $D$) if
$\rho$ is bound to $\pi$ in every context $\ell$ in which it occurs.

\begin{lemma}
  \label{lemma:completely-bound}
Let $C = \consOf{q}$.
Every region variable occurring in $\satC$ is bound to some region constant in $\satC$.
\end{lemma}

\begin{proof}
The constraint-generation rules guarantee that every region variable will
be unified with some region constant\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value

Note: We likely need to add another rule to the saturation rules for computing
$\satC$, namely $\isvalid{\ell}{r} \in \satC \Rightarrow \isvalid{\ell \conj \phi}{r} \in \satC$.
We need this to propagate constraints from an outer scope into an inner scope.

\TODO{Need to establish that there are no independent variables with same name.}
\end{proof}

\begin{lemma}
Let $C = \consOf{q}$.
Suppose a region variable $\rho$ is bound to two different region constants $\pi_1$
and $\pi_2$ in $\satC$.
Let $\soln$ satisfy $\satC$.
For any context $\varphi \conj \phi$ that contains $\rho$, we must have
$\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
$\isvalid{\predSubstFn(\varphi)}{\pi_1 = \pi_2}$.
\end{lemma}

\begin{proof}
  Assume that a region variable $\rho$ is bound to two different
  region constants $\pi_1$ and $\pi_2$ in a context $\varphi \conj \phi$.
  By transitivity, $\satC$ is guaranteed to have the constraints
  $\isvalid{\varphi \conj \phi}{\pi_1 \outlives \pi_2}$
  and $\isvalid{\varphi \conj \phi}{\pi_2 \outlives \pi_1}$.
  By abduction decomposition, one of the following conditions must hold:
  (a) $\{ \pi_1, \pi_2 \} \subseteq \predDeltaMap(\varphi)$ and
  both $\isvalid{\varphi}{\pi_1 \outlives \pi_2}$ and
  $\isvalid{\varphi}{\pi_2 \outlives \pi_1}$ are in $\satC$, or
  (b) $\{ \pi_1, \pi_2 \} \not\subseteq \predDeltaMap(\varphi)$ and
  both $\isvalid{\phi}{\pi_1 \outlives \pi_2}$ and
  $\isvalid{\phi}{\pi_2 \outlives \pi_1}$ are in $\satC$.
  In case (b), $\satC$ will not be satisfiable, so we can ignore this case.
  In case (a), $\varphi$ must be bound to a conjunction that includes both
  $\pi_1 \outlives \pi_2$ and $\pi_2 \outlives \pi_1$ in any satisfying assignment
  for $\satC$.
\end{proof}

\begin{lemma}
Let $C = \consOf{q}$.
Let $\satC_1$ denote the subset of $\satC$ obtained by removing all constraints
that contain a region variable.
$\satC$ is satisfiable iff $\satC_1$ is satisfiable.
\end{lemma}

\begin{proof}
  The forward implication is trivial.
  The reverse implication follows from Lemma~\label{lemma:completely-bound}, as shown below.

\end{proof}

\begin{theorem}
\label{thm:constraint-solver-soundness}
Let $C = \consOf{q}$.
If the constraint solver returns an assignment $\soln$,
then $\soln$ satisfies $\satC$.
\end{theorem}

\begin{proof}
Assume that the constraint solver returns an assignment $\soln$.

By construction, $\groundC$ (the set of all variable-free constraints in $\satC$)
is satisfiable. Hence, $\soln$ trivially satisfies all constraints in $\groundC$.

The definition of $\soln$ ensures that it satisfies the well-formedness constraints
for any region variable $\rho$.

Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in $\satC$,
where $\pi_i$ and $\pi_j$ are region constants. The definition of $\predSubstFn(\varphi)$
ensures that $\soln$ satisfies this constraint.
Furthermore, $\soln$ must also satisfy the well-formedness constraint
$\tywf{predDeltaMap(\varphi)}{\varphi}$.
(Otherwise, it follows from abduction decomposition that $\isvalid{}{\pi_i \outlives \pi_j}$ must
be in $\satC$, which means $\groundC$ must be invalid.)

Consider any constraint of the form $\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ in $\satC$,
It follows from abduction decomposition that either
$\isvalid{\varphi}{\pi_i \outlives \pi_j}$
or
$\isvalid{\phi}{\pi_i \outlives \pi_j}$ must be in $\satC$, which we know is satisfied by $\soln$.
Hence, $\soln$ must satisfy
$\isvalid{\varphi \conj \phi}{\pi_i \outlives \pi_j}$ as well.

This shows that $\soln$ satisfies all outlives-constraints in $\satC$ that have no variables in the consequent.

Consider any constraint of the form $\isvalid{\ell}{F_j(\varphi_j)}$.
Substituting the value $\predSubstFn(\varphi_j)$ reduces this constraint to one of
the form $\isvalid{\ell}{\conj_i \pi_i \outlives \pi_i'}$ where the substitution rule
guarantees $\isvalid{\ell}{\pi_i \outlives \pi_i'}$ is already in $\satC$ and, hence,
satisfied by $\soln$.
Hence, $\soln$ satisfies $\isvalid{\ell}{F_j(\varphi_j)}$ as well.

Consider any constraint of the form $\isvalid{\ell}{r}$ where $r$ contains a single occurrence
of a region variable, say $\rho$.
It follows from Lemma~\label{lemma:completely-bound} that $\rho$ is bound to some region
constant $\pi$.
Let $r'$ denote $r$ with $\rho$ replaced by $\pi$.
It follows from the transitivity rule that $\satC$ must contain the constraint $\isvalid{\ell}{r'}$
and that $\soln$ satisfies this constraint.
Further, it follows that $\soln$ satisfies $\isvalid{\ell}{\pi = \pi'}$ where $\pi' = \regionSubstFn(\rho)$.
Hence, $\soln$ satisfies this constraint as well.
\TODO{Flesh this out.}

The same idea works for constraints with occurrences of two region variables in the consequent.

\end{proof}

\begin{theorem}
\label{thm:constraint-solver-completeness}
Let $C = \consOf{q}$.
If $\satC$ is satisfiable, then the constraint solver will return some assignment.
\end{theorem}

\begin{proof}
If $\satC$ is satisfiable, then $\groundC$ is trivially satisfiable.
We just need to show that the definition $\regionSubstFn(\rho)$ for any
region variable $\rho$ will succeed.
\end{proof}

@@

Follows from Theorems~\ref{thm:constraint-solver-soundness} and~\ref{thm:constraint-solver-completeness}.

% Theorem~\ref{thm:closure} shows that $\satC$ and $C$ are equivalent.
% 
% The next step is to show that $\groundC$ and $\satC$ are (almost) equivalent.
% Consider $\satC \setminus \groundC$. This includes the well-formedness constraints.
% It is easy to show that we take care of these.
% 
% The more difficult case are validity constraints, involving either region variables
% ($\rho$) or predicate variables ($\varphi$) or both.
% 
% We use this property to show that it is okay to ignore validity constraints
% in $\satC$ containing region variables. For every
% region variable $\rho$ unified with a constant $\pi$, and for every validity
% constraint containing $\rho$, we will have a corresponding validity constraint
% containing $\pi$ instead.
% 
% As for validity constraints containing a predicate variable $\varphi$ on the consequent,
% we argue that the use of the Substitution rule in computing $\satC$ makes it okay
% to ignore these constraints.
% 
% As for validity constraints containing a predicate variable $\varphi$ on the antecedent,
% we use Lemma~\ref{lemma:decomposition} to argue that Abduction Decomposition lets
% us convert these into simpler constraints that do not have $\varphi$ on the
% antecedent or into a constraint that is directly incorporated into definition
% of $\predSubstFn$.
\end{proof}

@@

\begin{itemize}
  \item explain: non-determinism in defining region-variable binding is immaterial.
  \item explain lemma 2's use case (a) and (b)
  \item notation: decomposable is dependent on C
  \item cleanup: syntax of constraints (region-constraint, validity constraint, etc.)
    and later usage (abduction constraint)
  \item adjust spacing in RULEs
  \item for function-calls: need basic typing from core language to introduce region parameters
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item use of alpha-renaming, and its use to discharge not-in contraints;
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item update abstract/intro to stress aspects of constraint solver
  \item Change static semantics to use method precondition, without class precondition
\end{itemize}
